Comparison of Fast Imaging Architectures for Large, Multi-scale, Hierarchical Aperture Arrays

### 2024-06-15 ms.tex
Update on Overleaf.

\begin{abstract}
Two major science drivers of modern radio astronomy, namely, explosive astrophysical transient phenomena and observational cosmology, are pushing aperture arrays towards large-$N$, low-cost element design consisting of multiple spatial scales. Multi-scale aperture arrays require hybrid combinations of data processing architectures-- pre-correlation beamformer, generic version of FFT-based direct imager, post-correlation beamformer, and post-correlation FFT imager-- operating on different spatial scales to obtain optimal performance in imaging the entire field of view. We use the metric of computational cost density defined over the dimensions of discovery space, namely, field of view, angular resolution, polarisation, frequency, and time to determine the most efficient hybrid architecture for aperture arrays spanning multiple spatial scales like the element size, stations (a collection of elements), and the array (a collection of stations) with a range of number of elements per station and number of stations. We use nominal parameters of specific upcoming and planned arrays-- Square Kilometre Array (SKA) at low frequencies (\texttt{SKA-low}), \texttt{SKA-low-core}, a long baseline extension to SKA-low (\texttt{LAMBDA-I}), compact all-sky phased array (CASPA), and a lunar array (\texttt{FarView})-- to determine the most optimal architectural hierarchy from a computational standpoint, and provide a guide for designing hybrid architectures for multi-scale aperture arrays. For large-$N$, dense-packed layouts, a FFT-based direct imager is most efficient for most cadence intervals, and for other layouts that have relatively lesser number of elements or greater sparsity in distribution, the best architecture is more sensitive to the cadence interval.
\end{abstract}

@@

Based on the requirements for transients and observational cosmology, radio interferometers are witnessing a shift to a paradigm of packing a large number of collecting elements in a compact area with high density and large field of view. Modern radio arrays tend to be aperture arrays that operate interferometrically and hierarchically on multiple spatial scales. Examples of these multi-scale hierarchical aperture arrays include the Murchison Widefield Array \cite[MWA;][]{Tingay+2013}, the Hydrogen Epoch of Reionization Array \citep[HERA;][]{HERA+2017}, the Low Frequency Array \cite[LOFAR;][]{vanHaarlem+2013}, the swarm of Long Wavelength Arrays \cite[LWA Swarm;][]{Dowell+2018}, the Canadian Hydrogen Intensity Mapping Experiment \citep[CHIME;][]{CHIME+2022}, the Hydrogen Intensity and Real-time Analysis Experiment \citep[HIRAX;][]{HIRAX+2022}, and the Square Kilometre Array low \cite[SKA-low;][]{Dewdney+2009,SKA1+2019}.

This new paradigm poses severe computational and data rate challenges necessitated by the processing of large numbers of data streams in very short cadence intervals. The choice of data processing architecture is usually sensitive to the array layout, cadence, angular resolution, and field of view coverage. For example, a correlator-based architecture is generally more suited for relatively smaller number of interferometer elements sparsely distributed over a large area and processing on a slower cadence, whereas a direct imaging architecture based on Fast Fourier Transform (FFT) have far greater efficiency for regularly arranged elements \citep{Daishido+1991,Otobe+1994,Tegmark+2009,Tegmark+2010,Foster+2014,Masui+2019}. A generalised version of FFT-based direct imaging like the E-field Parallel Imaging ``Correlator'' \citep[EPIC;][]{Thyagarajan+2017,Thyagarajan+2019,Krishnan+2023} are found to be highly efficient for large-$N$, densely packed arrays with or without regularly placed elements from computational and data rate standpoints. Pre- or post-correlation beamforming is expected to be efficient when the number of elements and the number of locations where the image is sought are few in number. Thus, the choice of an optimal imaging architecture is critical for maximising the probability of discovery. When the chosen architecture is sub-optimal, compromises are often made by limiting one or more discovery parameters such as time resolution, on-sky observing time, bandwidth, field of view, or angular resolution \citep[for example,][]{Price2024}. When considering multi-scale arrays that involve hierarchical data processing at different scales, there may not a single architecture that is optimal at levels. A hybrid processing architecture that is optimal at various levels of data processing hierarchy is required.

In this paper, our primary motivation is to explore different imaging architectures and their combinations appropriate for a multi-dimensional space of array layout parameters and cadence intervals from a computational standpoint. Conversely, the paper also provides a guide to choosing the array layout parameters and cadence intervals where a given architecture or combinations thereof will remain computationally efficient. Several existing and hypothetical multi-scale aperture arrays are used as examples.

@@

The measuring instrument will imprint its transfer function on the measurements. If $a$ denotes a collecting element at location $\boldsymbol{r}_{a}$, then the electric field measured by this element will be
\begin{align}
    E_{a}^p &= \int_\mathbb{S} \sum_\alpha  \mathcal{W}_{a}^{p\alpha}(\hat{\boldsymbol{s}})\, \mathcal{E}^\alpha(\hat{\boldsymbol{s}})\, e^{-i \frac{2\pi}{\lambda} \hat{\boldsymbol{s}}\cdot\boldsymbol{r}_{a}} \,\mathrm{d}\Omega + n_{a}^p \, , \label{eqn:polarimetric-Efield-element-ideal}
\end{align}
where, $n_{a}^p$ and $\mathcal{W}_{a}^{p\alpha}(\hat{\boldsymbol{s}})$ are the stochastic noise and polarisation-sensitive directional electric field response of the element, $a$, respectively. $\mathcal{W}_{a}^{p\alpha}(\hat{\boldsymbol{s}})$ denotes the contribution from polarisation state, $\alpha$, in $\mathbb{S}$ to polarisation state, $p$, in the measurement.

Thus, the measured electric field can be treated as a linear superposition of the electric fields emanating from astronomical sources with appropriate complex phases acquired on account of propagation and properties of the measuring instrument. Ignoring wide-field effects, this relation can be simplified as a two-dimensional Fourier transform of the electric fields in the sky coordinates \cite{TMS2017}. Using the convolution theorem of Fourier transform, Equation~(\ref{eqn:polarimetric-Efield-element-ideal}) can be expressed as the weighted integral of the propagated electric field over the region the element is responsive to in the aperture surface, $\mathbb{R}$, 
\begin{align}
    E_{a}^p &= \int_\mathbb{R} W_{a}^{p\alpha}(\boldsymbol{r}_{a} - \boldsymbol{r})\,E^\alpha(\boldsymbol{r}) \, \mathrm{d}^2 \boldsymbol{r}+ n_{a}^p \, , \label{eqn:polarimetric-Efield-element-ideal-aperture-continuous}
\end{align}
where, the weighting, denoted by $W_{a}^{p\alpha}(\boldsymbol{r})$, is the spatial Fourier transform of $\mathcal{W}_{a}^{p\alpha}(\hat{\boldsymbol{s}})$,
\begin{align}
    W_{a}^{p\alpha}(\boldsymbol{r}) &= \frac{1}{\lambda^2}\int_\mathbb{S} \mathcal{W}_{a}^{p\alpha}(\hat{\boldsymbol{s}})\, e^{-i\frac{2\pi}{\lambda} \hat{\boldsymbol{s}}\cdot \boldsymbol{r}} \,\mathrm{d}\Omega \, . \label{eqn:holographic-Efield-illumination}
\end{align}

Equation~(\ref{eqn:polarimetric-Efield-element-ideal-aperture-continuous}) involves a weighted integral of the propagated electric fields, where the weighting corresponds to the instrumental response, $W_{a}^{p\alpha}(\boldsymbol{r})$, centered at the location of the element, $a$, in the aperture plane, $\mathbb{R}$. Generally, $W_{a}^{p\alpha}(\boldsymbol{r})$ is physically interpretable as the holographic pattern of the element element in the aperture plane, which has support over the element's area of influence where it collects signals from. 
% However, in some cases such as for narrow dipoles, this relation is purely mathematical and a physical interpretation is not straightforward. 
% such as narrow dipoles, this relation is  encapsulates the aperture field illumination pattern of the element or the element's aperture distribution, 
$W_{a_m}^{p\alpha}(\boldsymbol{r})$ can also include other instrumental effects that influence the element behavior.

Equations~(\ref{eqn:polarimetric-Efield-element-ideal}) and (\ref{eqn:polarimetric-Efield-element-ideal-aperture-continuous}) can be discretised as:
\begin{align}
    E_{a}^p &= \sum_k e^{-i\frac{2\pi}{\lambda}\hat{\boldsymbol{s}}_k\cdot\boldsymbol{r}_{a}} \, \sum_\alpha \left[\mathcal{W}_{a}^{p\alpha}(\hat{\boldsymbol{s}}_k) \, \mathcal{E}^\alpha(\hat{\boldsymbol{s}}_k)\right] \, \delta\Omega_k + n_{a}^p \, , \label{eqn:Efield-element-img-discrete}
\end{align}
and
\begin{align}
    E_{a}^p &= \sum_j \sum_\alpha W_{a}^{p\alpha}(\boldsymbol{r}_{a}-\boldsymbol{r}_j)\, \delta^2\boldsymbol{r}_j \left(\sum_k e^{-i\frac{2\pi}{\lambda} \hat{\boldsymbol{s}}_k\cdot\boldsymbol{r}_j} \, \mathcal{E}^\alpha(\hat{\boldsymbol{s}}_k) \,\delta\Omega_k \right) \nonumber\\ 
    &\quad + n_{a}^p \, , \label{eqn:Efield-element-aperture-discrete}    
\end{align}
respectively. Equation~(\ref{eqn:Efield-element-aperture-discrete}) can be represented linear algebraically as
\begin{align}
    \mathbf{E}_{[p,\boldsymbol{r}_{a}]} &= \mathbf{W}_{[p,\boldsymbol{r}_{a}\leftarrow \alpha,\boldsymbol{r}]} \, \mathbf{F}_{[\boldsymbol{r}\leftarrow \hat{\boldsymbol{s}}]} \, \boldsymbol{\mathcal{E}}_{[\alpha,\hat{\boldsymbol{s}}]} + \mathbf{n}_{[p,\boldsymbol{r}_{a}]} \, .
\end{align}
The subscripted variables on either side of $\leftarrow$ indicate the transformation carried out by the associated matrix. For example, $\mathbf{F}_{[\boldsymbol{r}\leftarrow \hat{\boldsymbol{s}}]}\coloneqq e^{-i\frac{2\pi}{\lambda} \hat{\boldsymbol{s}}_k\cdot\boldsymbol{r}_j}$ denotes the Fourier matrix transforming variables defined on sky grid, $\hat{\boldsymbol{s}}_k$, to the aperture grid, $\boldsymbol{r}_j$. And, $\mathbf{W}_{[p,\boldsymbol{r}_{a}\leftarrow \alpha,\boldsymbol{r}]}\coloneqq \{W_{a}^{p\alpha}(\boldsymbol{r}_{a}-\boldsymbol{r}_j)\}$ is a matrix, where each row corresponds to the holographic pattern of an element, $W_{a}^{p\alpha}(\boldsymbol{r}_j)$, and the columns correspond to the aperture grid locations, $\boldsymbol{r}_j$. In each row, the holographic weights are shifted to the element location, $\boldsymbol{r}_{a}$.

@@

The hierarchical, multi-scale aperture array scenario we consider involves spatial scales corresponding to the dimensions of the antenna, station (a grouping of antennas), and the array (a grouping of stations), denoted by $D_\textrm{a}$, $D_\textrm{s}$, and $D_\textrm{A}$, respectively. The number of stations in the array and the number of antennas in a station (indexed by $m$) are denoted by $N_\textrm{s}$ and $N_{\textrm{a/s}_m}$, respectively. 
% Although $N_\textrm{a/s_m}$ could vary with each station, we assume it remains the same for each station for simplicity.

@@

Although alternate computational metrics that include additional components, such as memory bandwidth, could be justified from a practical standpoint, these metrics will vary significantly with real-life implementations and hardware components available. Therefore, we take a theoretical approach with the justification that the budget of floating point operations will form a lower bound regardless of the details of any particular implementation.

>>>>

\begin{abstract}
Two major science drivers of modern radio astronomy, namely, explosive astrophysical transient phenomena and observational cosmology, are pushing aperture arrays towards large-$N$, low-cost element design consisting of multiple spatial scales. Multi-scale, hierarchical aperture arrays require a combination of data processing architectures-- pre-correlation beamformer, generic version of FFT-based direct imager, post-correlation beamformer, and post-correlation FFT imager-- operating on different spatial scales to obtain optimal performance in imaging the entire field of view. We use the metric of computational cost density defined over the dimensions of discovery space, namely, field of view, angular resolution, polarisation, frequency, and time to determine the most efficient hybrid architecture for aperture arrays spanning multiple spatial scales like the element size, stations (a collection of elements), and the array (a collection of stations) with a range of number of elements per station and number of stations. We use nominal parameters of specific upcoming and planned arrays-- Square Kilometre Array (SKA) at low frequencies (\texttt{SKA-low}), \texttt{SKA-low-core}, a long baseline extension to SKA-low (\texttt{LAMBDA-I}), compact all-sky phased array (CASPA), and a lunar array (\texttt{FarView})-- to determine the most optimal architectural hierarchy from a computational standpoint, and provide a guide for designing hybrid architectures for multi-scale aperture arrays. For large-$N$, dense-packed layouts, a FFT-based direct imager is most efficient for most cadence intervals, and for other layouts that have relatively lesser number of elements or greater sparsity in distribution, the best architecture is more sensitive to the cadence interval.
\end{abstract}

@@

Based on the requirements for transients and observational cosmology, radio interferometers are witnessing a paradigm shift towards packing a large number of collecting elements densely in a compact region providing a large field of view. Modern radio arrays tend to be aperture arrays that operate interferometrically and hierarchically on multiple spatial scales. Examples of multi-scale, hierarchical aperture arrays include the Murchison Widefield Array \cite[MWA;][]{Tingay+2013}, the Hydrogen Epoch of Reionization Array \citep[HERA;][]{HERA+2017}, the Low Frequency Array \cite[LOFAR;][]{vanHaarlem+2013}, the swarm of Long Wavelength Arrays \cite[LWA Swarm;][]{Dowell+2018}, the Canadian Hydrogen Intensity Mapping Experiment \citep[CHIME;][]{CHIME+2022}, the Hydrogen Intensity and Real-time Analysis Experiment \citep[HIRAX;][]{HIRAX+2022}, and the Square Kilometre Array low \cite[SKA-low;][]{Dewdney+2009,SKA1+2019}.

This new paradigm poses severe computational and data rate challenges necessitated by the processing of large numbers of data streams in very short cadence intervals. The choice of data processing architecture is usually sensitive to the array layout, cadence, angular resolution, and field of view coverage. For example, a correlator-based architecture is generally more suited for relatively smaller number of interferometer elements sparsely distributed over a large area and processing on a slower cadence, whereas a direct imaging architecture based on Fast Fourier Transform (FFT) have far greater efficiency for regularly arranged elements \citep{Daishido+1991,Otobe+1994,Tegmark+2009,Tegmark+2010,Foster+2014,Masui+2019}. A generalised version of FFT-based direct imaging like the E-field Parallel Imaging ``Correlator'' \citep[EPIC;][]{Thyagarajan+2017,Thyagarajan+2019,Krishnan+2023} are found to be highly efficient for large-$N$, densely packed arrays with or without regularly placed elements from computational and data rate standpoints. Pre- or post-correlation beamforming is expected to be efficient when the number of elements and the number of locations where the image is sought are few in number. Thus, the choice of an optimal imaging architecture is critical for maximising the discovery potential. When the chosen architecture is sub-optimal, compromises are often made by limiting one or more discovery parameters such as time resolution, on-sky observing time, bandwidth, field of view, or angular resolution \citep[for example,][]{Price2024}. When considering multi-scale arrays that involve hierarchical data processing, there may not a single architecture that is optimal on all scales. A hybrid processing architecture that is optimal at various levels of data processing hierarchy is required.

In this paper, our primary motivation is to explore different imaging architectures and their combinations appropriate for aperture arrays spanning a multi-dimensional space of array layout parameters and cadence intervals from a computational standpoint. Conversely, the paper also provides a guide to choosing the array layout parameters and cadence intervals where a given architecture, or combinations thereof, will remain computationally efficient. Several upcoming and hypothetical multi-scale aperture arrays are used as examples.

@@

The measuring instrument will imprint its transfer function on the measurements. If $a$ denotes a collecting element at location $\boldsymbol{r}_{a}$, then the electric field measured by this element will be
\begin{align}
    E_{a}^p &= \int_\mathbb{S} \sum_\alpha \mathcal{W}_{a}^{p\alpha}(\hat{\boldsymbol{s}})\, \mathcal{E}^\alpha(\hat{\boldsymbol{s}})\, e^{-i \frac{2\pi}{\lambda} \hat{\boldsymbol{s}}\cdot\boldsymbol{r}_{a}} \,\mathrm{d}\Omega + n_{a}^p \, , \label{eqn:polarimetric-Efield-element-ideal}
\end{align}
where, $n_{a}^p$ and $\mathcal{W}_{a}^{p\alpha}(\hat{\boldsymbol{s}})$ are the stochastic noise and polarisation-sensitive, directional electric field response of the element, $a$, respectively. $\mathcal{W}_{a}^{p\alpha}(\hat{\boldsymbol{s}})$ denotes the contribution from polarisation state, $\alpha$, in $\mathbb{S}$ to polarisation state, $p$, in the measurement.

Thus, the measured electric field can be treated as a linear superposition of the electric fields emanating from astronomical sources with appropriate complex phases acquired on account of propagation and properties of the measuring instrument. Ignoring wide-field effects, this relation can be simplified as a two-dimensional Fourier transform of the electric fields in the sky coordinates \citep{TMS2017}. Using the convolution theorem of Fourier transform, Equation~(\ref{eqn:polarimetric-Efield-element-ideal}) can be expressed as the weighted integral of the propagated electric field over the region the element is responsive to in the aperture surface, $\mathbb{R}$, 
\begin{align}
    E_{a}^p &= \int_\mathbb{R} \sum_\alpha W_{a}^{p\alpha}(\boldsymbol{r}_{a} - \boldsymbol{r})\,E^\alpha(\boldsymbol{r}) \, \mathrm{d}^2 \boldsymbol{r}+ n_{a}^p \, , \label{eqn:polarimetric-Efield-element-ideal-aperture-continuous}
\end{align}
where, the weighting, denoted by $W_{a}^{p\alpha}(\boldsymbol{r})$, is the spatial Fourier transform of $\mathcal{W}_{a}^{p\alpha}(\hat{\boldsymbol{s}})$,
\begin{align}
    W_{a}^{p\alpha}(\boldsymbol{r}) &= \frac{1}{\lambda^2}\int_\mathbb{S} \mathcal{W}_{a}^{p\alpha}(\hat{\boldsymbol{s}})\, e^{-i\frac{2\pi}{\lambda} \hat{\boldsymbol{s}}\cdot \boldsymbol{r}} \,\mathrm{d}\Omega \, . \label{eqn:holographic-Efield-illumination}
\end{align}

Equation~(\ref{eqn:polarimetric-Efield-element-ideal-aperture-continuous}) involves a weighted integral of the propagated electric fields, where the weighting corresponds to the instrumental response, $W_{a}^{p\alpha}(\boldsymbol{r})$, centred at the location of the element, $a$, in the aperture plane, $\mathbb{R}$. Generally, $W_{a}^{p\alpha}(\boldsymbol{r})$ is physically interpretable as the holographic pattern of the element element in the aperture plane, which has support over the element's area of influence where it collects signals from. 
% However, in some cases such as for narrow dipoles, this relation is purely mathematical and a physical interpretation is not straightforward. 
% such as narrow dipoles, this relation is  encapsulates the aperture field illumination pattern of the element or the element's aperture distribution, 
$W_{a}^{p\alpha}(\boldsymbol{r})$ can also include other instrumental effects that influence the element behavior.

Equations~(\ref{eqn:polarimetric-Efield-element-ideal}) and (\ref{eqn:polarimetric-Efield-element-ideal-aperture-continuous}) can be discretised as:
\begin{align}
    E_{a}^p &= \sum_k e^{-i\frac{2\pi}{\lambda}\hat{\boldsymbol{s}}_k\cdot\boldsymbol{r}_{a}} \, \sum_\alpha \left[\mathcal{W}_{a}^{p\alpha}(\hat{\boldsymbol{s}}_k) \, \mathcal{E}^\alpha(\hat{\boldsymbol{s}}_k)\right] \, \delta\Omega_k + n_{a}^p \, , \label{eqn:Efield-element-img-discrete}
\end{align}
or
\begin{align}
    E_{a}^p &= \sum_j \sum_\alpha W_{a}^{p\alpha}(\boldsymbol{r}_{a}-\boldsymbol{r}_j)\, \delta^2\boldsymbol{r}_j \left(\sum_k e^{-i\frac{2\pi}{\lambda} \hat{\boldsymbol{s}}_k\cdot\boldsymbol{r}_j} \, \mathcal{E}^\alpha(\hat{\boldsymbol{s}}_k) \,\delta\Omega_k \right) \nonumber\\ 
    &\quad + n_{a}^p \, , \label{eqn:Efield-element-aperture-discrete}    
\end{align}
respectively. Equation~(\ref{eqn:Efield-element-aperture-discrete}) can be represented linear algebraically as
\begin{align}
    \mathbf{E}_{[p,\boldsymbol{r}_{a}]} &= \mathbf{W}_{[p,\boldsymbol{r}_{a}\leftarrow \alpha,\boldsymbol{r}]} \, \mathbf{F}_{[\boldsymbol{r}\leftarrow \hat{\boldsymbol{s}}]} \, \boldsymbol{\mathcal{E}}_{[\alpha,\hat{\boldsymbol{s}}]} + \mathbf{n}_{[p,\boldsymbol{r}_{a}]} \, .
\end{align}
The subscripted variables on either side of $\leftarrow$ indicate the transformation carried out by the associated matrix. For example, $\mathbf{F}_{[\boldsymbol{r}\leftarrow \hat{\boldsymbol{s}}]}\coloneqq e^{-i\frac{2\pi}{\lambda} \hat{\boldsymbol{s}}_k\cdot\boldsymbol{r}_j}$ denotes the Fourier matrix transforming variables defined on sky grid, $\hat{\boldsymbol{s}}_k$, to the aperture grid, $\boldsymbol{r}_j$. And, $\mathbf{W}_{[p,\boldsymbol{r}_{a}\leftarrow \alpha,\boldsymbol{r}]}\coloneqq \{W_{a}^{p\alpha}(\boldsymbol{r}_{a}-\boldsymbol{r}_j)\}$ is a matrix, where each row corresponds to the holographic pattern of an element, $W_{a}^{p\alpha}(\boldsymbol{r}_j)$, and the columns correspond to the aperture grid locations, $\boldsymbol{r}_j$. In each row, the holographic weights are shifted to the element location, $\boldsymbol{r}_{a}$.

@@

The hierarchical, multi-scale aperture array scenario we consider involves spatial scales corresponding to the dimensions of the collecting element, station (a grouping of elements), and the array (a grouping of stations), denoted by $D_\textrm{a}$, $D_\textrm{s}$, and $D_\textrm{A}$, respectively. The number of stations in the array and the number of elements in a station (indexed by $m$) are denoted by $N_\textrm{s}$ and $N_{\textrm{a/s}_m}$, respectively. 
% Although $N_\textrm{a/s_m}$ could vary with each station, we assume it remains the same for each station for simplicity.

@@

Although alternate computational metrics that include additional components, such as memory bandwidth, could be justified from a practical standpoint, these metrics will vary significantly with real-life implementations and hardware components available during deployment. Therefore, we take a theoretical approach with the justification that the budget of floating point operations will form a lower bound immune to the details of any particular implementation.
